---
title: "Dates"
layout: default
output: bookdown::html_chapter
---

# Dates

A standard historian's joke goes something like this: "On or about October 31, 1517, modernity began." Martin Luther did indeed nail his ninety-five theses to the door of the Wittenberg church on that date. Identifying the beginning of modernity or capitalism, or conceptions of the self, or any other historical abstract worth studying cannot be so readily pinned down to a specific date, whether on or about. The problem is one of precision. Historical events and movements have fuzzy beginnings and endings and often even straightforward facts often have uncertain dates. Think of chronological terms like "the long nineteenth century," "the last third of the eighteenth century," "the American Revolution," or "antiquity" and "modernity," and you'll recognize how good historians are expressing the necessary chronological uncertainty in prose.

The problem of precision and dates is even more vexing when it comes to computation. Computers demand precision in dates. The computer has no built-in concept of dates such as centuries. To explain the nineteenth century to a computer, you will have to give tell it to computer the difference between December 31, 1900 and January 1, 1801.^[Or 1800 and 1899: whichever definition of the start of a century you prefer.] Even more difficult is that computers require precision down to the second or below. On Unix operating systems, dates are computed as the number of seconds elapsed since the beginning of the Unix epoch: 00:00:00 in Coordinated Universal Time (i.e., Greenich Mean Time) on Thursday, January 1, 1970, excluding leap seconds. If I run the Unix date command, I see that 1.4 billion seconds and counting have elapsed since then, and R reports a value down to the hundredth thousand of a second.

```{r get_sys_time}
now <- Sys.time()
as.character(as.numeric(now))
```

The problem for the computational historian, then, is to work with artificially precise dates in a way that makes sense for the discipline of history. 

## Lubridate

```{r demo_lubridate}
library(lubridate)
date <- mdy("August 8, 1863")
date
class(date)
```

## Parsing large quantities of dates
