# Getting started

This chapter will give you the minimum information you need to know in order to get started reading this book, and it will point you to other places in the book where you can find more information. 

## Installing R and RStudio

R is freely available from [CRAN](https://cran.r-project.org/) (the Comprehensive R Archive Network). How you install R will depend on your operating system. See the sections on installing R on a Mac (\@ref(install-mac)), Windows (\@ref(install-windows)), or Ubuntu (\@ref(install-ubuntu)) for details.

You should also install [RStudio Desktop](https://www.rstudio.com/products/rstudio/), an integrated development environment for R. While RStudio is not strictly necessary, you will find it enormously helpful for writing R code and understanding what is happening in your session.

## Packages you will need

Most of this book uses the [tidyverse](http://tidyverse.org/) of R packages. These packages are designed to work together using a concept called tidy data. For an explanation of this concept, see Hadley Wickham's "[The Tidy Tools Manifesto](https://cran.rstudio.com/web/packages/tidyverse/vignettes/manifesto.html)," which is part of the tidyverse package.^[@R-tidyverse. For a more rigorous explanation, see @wickham_tidy_2014.] We can install most of the packages that we need by installing the [tidyverse](https://cran.r-project.org/package=tidyverse) package. For now, you can run the following code in your R session. See the section on installing packages from CRAN for a fuller explanation (section \@ref(install-cran)). 

```{r, eval=FALSE}
install.packages("tidyverse")
```

Most of the examples in this book are available in the [historydata](https://cran.r-project.org/package=historydata) package. While this book is being written, you will need the development version of the package, which is [available on GitHub](https://github.com/ropensci/historydata/#readme). You can install it directly from GitHub using the [devtools](https://cran.r-project.org/package=devtools) package by running the code below. See the section on installing packages from GitHub for a fuller explanation (section \@ref(install-github)).

```{r, eval=FALSE}
install.packages("devtools")
devtools::install_github("ropensci/historydata")
```

You can test that you have these packages installed by loading them in your R session. Note that tidyverse prints a message about the other packages that it is loading, but this is not an error message. See the section on packages and libraries for more information about loading packages (section \@ref(packages-libraries)).

```{r}
library(historydata)
library(tidyverse)
```

You will need other packages for different chapters in this book. You can install them from CRAN as you go. See the package guide for a description of other packages that may be helpful to you (appendix \@ref(package-guide)).
