# Installing R and packages

## What is CRAN?

R has very good support on the three main platforms: Linux, MacOS, and Windows. Both R and the language and the thousands of published packages which extend its functionality for the most part work across the three platforms. In addition, installing packages in R can be considerably simpler than in other languages. Much of the ease of installing R and add-on packages is thanks to the CRAN infrastructure.

R and the published packages are available at [CRAN](https://cran.r-project.org/), the "Comprehensive R Archive Network." The CRAN maintainers rigorously screen new packages, then test all accepted packages continually. This does not mean that there are no bugs in published R packages, but it does mean that the quality of software published at CRAN is generally higher than in the software repositories of other programming languages.

CRAN is a collection of mirrors, that is, servers that contain the exact same content.  The canonical URL for CRAN is `https://cran.r-project.org/`, but you can also access a CRAN mirror at Indiana University, for example, at `https://ftp.ussg.iu.edu/CRAN/`. The R Project maintains a [current list of mirrors](https://cran.r-project.org/mirrors.html). When installing packages at the start of an R session, R will ask you which mirror you want to use. Unless you have a compelling reason to pick a different one, you should select the first option, "Cloud": `https://cloud.r-project.org/`. That option will pick a server which is geographically close to you. If you add the following line to the `.Rprofile` file in your user's home directory (`~/.Rprofile` on Mac and Linux), then you can set that option permanently.

```{r eval=FALSE}
options(repos = c(CRAN = "https://cloud.r-project.org"))
```

## Installing R on MacOS

CRAN compiles R for the Mac. To install R, go to [CRAN](https://cran.r-project.org/) and click on the link to [download R for Mac](https://cran.r-project.org/bin/macosx/). Download the installer for the current version of R, for example, `R-3.3.2.pkg`. Install this package like you would any other Mac software. The installer will install R, plus various other software such as a GUI application for R.

You can test that you have R installed properly by opening the Terminal program on your Mac.

```
R --version
```

If you get a message that prints the current version of R, then you have successfully installed R.

In general, you should use the version of R that CRAN compiled for the Mac unless you have a specific reason not to do so. One of the chief advantages of this version is that you can install R extension packages which have been pre-compiled by CRAN for the Mac. This can make it significantly easier to install R packages.

You can however install R via Homebrew. [Homebrew](http://brew.sh/) is a package manager for the Mac which lets you install and update software from the command line. You might want to install R this way if you plan on compiling your own R packages, or want to try to optimize the linear algebra library which lets R perform matrix operations quickly.

First follow the instructions to install [Homebrew](http://brew.sh/) on its website. Then in your terminal, run the following commands. These add a list of science-related packages to the list of software that Homebrew knows how to install, and installs R. This will probably involve installing a lot of additional software, such as compilers.^[For a fuller explanation of installing R via Homebrew, see [Bob Rudis's post](http://rud.is/b/2015/10/22/installing-r-on-os-x-100-homebrew-edition/) on the topic.]

```
brew tap homebrew/science
brew install R
```

You can see the various options for installing R via Homebrew by running `brew show R` in the terminal. For instance, using certain flags you can change the linear algebra system that R is compiled against. If you install R with Homebrew, you may also need to install various system libraries, such as `libxml2` or `libcurl`.

## Installing R on Ubuntu
 Being able to install R on a Linux system is useful in several ways. Of course, if your primary computer runs a Linux distribution, then you can use R on it like you would on Mac or Windows. But you can also install R on a Linux server, or run R on a Linux virtual machine hosted on a different operating system. These instructions give you the commands for installing R on either the server or desktop version of Ubuntu 16.04.x, the latest long-term support (LTS) version of a popular Linux distribution.^[For a fuller explanation of how to install R on Ubuntu, see the [official instructions on CRAN](https://cran.r-project.org/bin/linux/ubuntu/README.html) or [Digital Ocean's guide](https://www.digitalocean.com/community/tutorials/how-to-install-r-on-ubuntu-16-04-2).]

Software on Ubuntu is generally installed via the apt package manager. A version of R is available in the Ubuntu repositories, but it it almost certainly not the most recent versions, especially in LTS versions of Ubuntu. Therefore it is best to add a CRAN-maintained Ubuntu repository to the list of software sources that apt checks, then install R.

First, run this command (and all following commands) in the Ubuntu terminal. This command gets the security key to verify the software.

```
sudo apt-key adv --keyserver keyserver.ubuntu.com --recv-keys E298A3A825C0D65DFD57CBB651716619E084DAB9
```

Then add the CRAN maintained Ubuntu repository.

```
sudo add-apt-repository 'deb https://cran.rstudio.com/bin/linux/ubuntu xenial/'
```

Finally install R. In addition to installing R, these commands will install the Ubuntu packages that you need to be able to compile R extension packages for yourself.

```
sudo apt update
sudo apt install r-base r-base-dev
```

If you can run the following command and get the correct version of R, then you have installed R correctly.

```
R --version
```

## Installing R on Windows

CRAN maintains a version of R for Windows, including a GUI, along with helpful documentation on how to install it. Follow the instructons on the [CRAN download page for Windows](https://cran.r-project.org/bin/windows/base/).

The version of R that you install from CRAN for Windows will be able to install binary (i.e., precompiled) versions of R extension packages. This is considerably easier than trying to compile those packages yourself. However, CRAN also maintains a set of [Rtools for Windows](https://cran.r-project.org/bin/windows/Rtools/) which provide the software that you will need to compile packages for yourself. If you wish to build packages from source, then install the version of Rtools that corresponds to the version of R you have installed.

## About R packages and libraries

CRAN hosts many packages for R. These packages extend R's functionality by providing additional functions. And R packages can also include data files, which is an easy way to distribute example data or data for learning that is not too large.^[For an explanation of how R packages are created, see @wickham_r_2015.]

These extensions to R are called **packages**. In other programming languages they might be called libraries. But in R, the term **library** is used for the directories where R packages are installed. It is important to remember that R can load packages from multiple libraries at a time. You can see which libraries your R session knows about with the function `.libPaths()`.

```{r}
.libPaths()
```

The second library in that list of directories is the library that comes bundled with R. It includes packages like the stats and utils packages which provide basic functionality to R. This library is kept in a system directory and is available to all users of R on a system.

The first library in that list is a personal package library. It is kept inside the user's home directory, in this case, the user `lmullen`. This library contains packages which the user has installed, and they are only accessible to that user. When you load a package with `library()` or `require()`, those functions look for a package in the first directory returned by `.libPaths()`, then the second and so on, stopping when it finds the first match. This ensures that versions of packages that you install are loaded in preference to system libraries. When installing packages for the first time, you will probably be prompted to create a personal package library.^[There are other places that R libaries can be kept. For example, the [packrat](https://cran.r-project.org/package=packrat) package can add a library to a project directory.]

## Installing packages from CRAN

![Page for the [dplyr](https://cran.r-project.org/package=dplyr) package on CRAN.](images/dplyr-cran-listing.png)

## References
