---
title: "Mapping"
layout: default
output: bookdown::html_chapter
---

```{r,echo=FALSE}
source("general-options.r")
```

# Mapping

Mapping is difficult because of the difficulty in representing geospatial data. In particular, the dominant format for spatial data is the [shapefile](http://en.wikipedia.org/wiki/Shapefile), which is a proprietary but mostly open format with some severe limitations. Geospatial data can be very large: in the hundreds of megabytes. Often it is unnecessarily large for the resolution of the map that you want to make, and must be simplified in order to be manageable. Finally, geospatial information is projected, meaning that it can represent a point's position on the earth by a latitude or longitude but also by any other number of [coordinate reference systems](http://en.wikipedia.org/wiki/Spatial_reference_system). These systems have to be translated between one another. In general, it is easiest to prepare geospatial data outside of R and then to load and import it. This chapter focuses primarily on what you can do with R, but the appendix demonstrates how to take a shapefile and then convert and simplify it into a useable format.

Another problem with geospatial information is that it is usually intended for contemporary use, which means that the information about boundaries and even about natural features like rivers and coastlines may be anachronistic. You will have to be cautious about finding geospatial information that fits your time and place.

There are some good resources for getting geospatial data. [NHGIS](http://nhgis.org) contains aggregated census information for the United States keyed by special codes to shapefiles that they provide. We will use NHGIS data about manufacturing in this chapter.[Natural Earth](http://www.naturalearthdata.com/) provides all kinds of geographic information, such as rivers, coastlines, and mountains.

Nevertheless, mapping is the digital history technique from which it is easiest to make historical arguments. Not coincidentally, it is also the digital history technique which is easiest to explain to non-digital historians. Historians have been working with maps since at least the beginnings of the historical profession: digital mapping gives you the capability to map data and places yourself.

## Installation and setup

Most R packages that work with geospatial data are wrappers around libraries written in lower-level languages. In order to use these packages, you will have to install those libraries which are external to R. The two main libraries are [GDAL/OGR](http://www.gdal.org/), which you will likely have to install anyway in order to get the command line tools to prepare your geospatial data, and [GEOS](http://trac.osgeo.org/geos/).^[Make sure you get version 1.11.0 or later of GDAL/OGR, which added support for [topojson](https://github.com/mbostock/topojson).] (In most cases, the R wrapper for GEOS can install the external library itself.)

You can install these packages by following the instructions on their respective websites. If you are on Mac and use [Homebrew](http://brew.sh/), you can install them with the following commands.

```
brew update
brew install gdal
brew install geos
```

If you are on Ubuntu, you can install them with the following commands. These libraries will be found in almost any Linux package manager.

```
apt-get update
apt-get install gdal-bin libgdal-dev libproj-dev
apt-get install libgeos-dev
```

Once you have those libraries installed, you can install the necessary R packages, which include the following:

- [sp](http://cran.r-project.org/web/packages/sp/index.html) provides the classes which describe geospatial data, and it will let you use base R functions like `plot()` to make maps. Virtually every geospatial package for R depends on this package.
- [rgdal](http://cran.r-project.org/web/packages/rgdal/index.html) provides functions to read and write spatial datasets.
- [rgeos](http://cran.r-project.org/web/packages/rgeos/index.html) provides functions for manipulating spatial information.
- [maptools](http://cran.r-project.org/web/packages/maptools/index.html) provides functions for manipulating spatial information.
- [ggmap](http://cran.r-project.org/web/packages/ggmap/ggmap.pdf) lets you map spatial information on top of Google Maps, but its chief usefulness is for its `geocode()` function which uses Google's API.
- [ggplot2](http://cran.r-project.org/web/packages/ggplot2/index.html) will let us plot spatial information using the same grammer we used for [plotting](plotting.html).

You can install all of these with the following command.

```{r eval=FALSE}
install.packages(c("sp", "rgdal", "rgeos", "maptools",
                   "ggmap", "ggplot2"),
                 dependencies = TRUE)
```

## Loading and plotting a shapefile

## Other stuff

```{r}
library(rgdal)
library(ggplot2)
library(dplyr)
counties_1850 <- readOGR("data/nhgis-employment/out", "US_county_1850")
map <- fortify(counties_1850, region = "GISJOIN")
employment <- read.csv("data/nhgis-employment/nhgis0030_csv/nhgis0030_ds10_1850_county.csv", stringsAsFactors = FALSE)
map <- map %>% left_join(employment, by = c("id" = "GISJOIN")) 
ggplot() + geom_map(data = map, map = map, aes(map_id = id, x = long, y = lat, group = group, fill = cut(AER001, breaks = quantile(map$AER001, na.rm =TRUE))), color = "white", size = 0.05) + coord_map(projection = "polyconic") + theme_minimal() + theme(legend.position="bottom")
```


## Geocoding

Given a set of names or addresses, how to get latitudes and longitudes for them.

## Named Entity Extraction

How to get place (and other kinds of names) out of a document.

## Appendix: Preparing Geospatial Data


## Next Steps

Further Reading

- Bivand, Roger S., Edzer Pebesma, and Virgilio Gómez-Rubio. *[Applied Spatial Data Analysis with R](http://link.springer.com/book/10.1007/978-1-4614-7618-4)*. Springer, 2013.
- Bodenhamer, David J., John Corrigan, and Trevor M. Harris. *The Spatial Humanities: GIS and the Future of Humanities Scholarship*. Bloomington: Indiana University Press, 2010.
- Gregory, Ian N. and Alistair Geddes. *Toward Spatial Humanities: Historical GIS and Spatial History*. Bloomington: Indiana University Press, 2014.
