# An R primer {#primer}

What follows is a highly condensed guide to the basic features of the R programming language and some of its most useful general purpose packages. Use this primer as a quick introduction to the language, or as a reference when reading the rest of the book. 

## Values

A **value** is the most basic piece of data. There are several kinds of values.

**Doubles** are positive or negative numbers that can have a decimal point. **Integers** are positive or negative whole numbers. Even if a value does not have a decimal point, it is a double unless it is suffixed with `L`. 

```{r}
10.42
-8.67
10L
-8L
```

Text is stored in **character** vectors (usually called strings in other programming langauges). Characters are marked by surrounding single or double quotation marks.

```{r}
"Everything is awesome"
"1" # not the same as 1
identical(1, "1")
```

**Logical** values, or booleans, can be either true or false.

```{r}
TRUE
FALSE
```

**Factors** are a special kind of text value, where the values must be part of a predefined set of options. Factors are usually used for categorical data. 

```{r}
factor(c("correct", "correct", "incorrect"), levels = c("correct", "incorrect"))
```

Each of these kinds of values has a corresponding missing value, `NA`. These are used for missing or unrecorded data.

```{r}
NA
NA_character_
NA_integer_
NA_real_
```

## Variables and assignment

Values can be stored in **variables**. A value is placed into a variable using the assignment operator, `<-`.

```{r}
x <- 1
x
y <- -3.14
y
result <- "The results are positive"
result
```

## Comparison

Values can be compared to one another. The result is always a vector of `TRUE` and `FALSE` values indicating whether the condition was met or not. 

```{r}
x <- 1:6 # a vector of numbers 1 to 6
x > 3
x >= 3
x < 3
x == 3
x != 3
```

Comparison also works with character vectors.

```{r}
"Is this the same?" == "as this?"
"Is this the same?" == "Is this the same?"
```

Notice that equality is tested with the `==` operater, not with `=`.

Often it is helpful to test whether elements of one vector are contained in another vector with the `%in%` function. Notice that the resulting vector is the same length as the vector on the left-hand side.

```{r}
c(1, 4, 2, 8, 0, 10) %in% c(1, 2, 3)

# Are these states a part of New England?
c("VA", "CT", "MA", "SD", "GA", "AL") %in% c("MA", "ME", "CT", "RI", "NH", "VT")
```


