# An R primer

This book is not intended to teach you how to program in R. If you want a general purpose introduction to the language, I recommend the following sequence of books.

- Garret Grolemund's *Hands-On Programming with R* is a good introduction to the basics of R programming.^[@grolemund_hands-programming_2014.]
- Hadley Wickham and Garret Grolemund's [*R for Data Science*](http://r4ds.had.co.nz/) teaches some of the basics of R programming in the process of teaching data analysis. If you have already programmed before, this is the place to start.^[@wickham_r_2017.]
- Hadley Wickham's [*Advanced R*](http://adv-r.had.co.nz/) is the definitive guide to the language. Once you have a journeyman's proficiency in R, use this book to gain a deeper knowledge.^[@wickham_advanced_2014.]
- Norman Matloff's *The Art of R Programming* is another excellent, in-depth guide to the language as a whole.^[@matloff_art_2011.]

What follows is a highly condensed guide to the basic features of the R programming language and some of its most useful general purpose packages. Use this primer as a quick introduction to the language, or as a reference when reading the rest of the book. 

## Values

A **value** is the most basic piece of data. There are several kinds of values.

**Doubles** are positive or negative numbers that can have a decimal point. **Integers** are positive or negative whole numbers. Even if a value does not have a decimal point, it is a double unless it is suffixed with `L`. 

```{r}
10.42
-8.67
10L
-8L
```

Text is stored in **character** vectors (usually called strings in other programming langauges). Characters are marked by surrounding single or double quotation marks.

```{r}
"Everything is awesome"
"1" # not the same as 1
identical(1, "1")
```

**Logical** values, or booleans, can be either true or false.

```{r}
TRUE
FALSE
```

**Factors** are a special kind of text value, where the values must be part of a predefined set of options. Factors are usually used for categorical data. 

```{r}
factor(c("correct", "correct", "incorrect"), levels = c("correct", "incorrect"))
```

Each of these kinds of values has a corresponding missing value, `NA`. These are used for missing or unrecorded data.

```{r}
NA
NA_character_
NA_integer_
NA_real_
```
