# An R primer {#primer}

What follows is a highly condensed guide to the basic features of the R programming language and some of its most useful general purpose packages. Use this primer as a quick introduction to the language, or as a reference when reading the rest of the book. 

## Values

A **value** is the most basic piece of data. There are several kinds of values.

**Doubles** are positive or negative numbers that can have a decimal point. **Integers** are positive or negative whole numbers. Even if a value does not have a decimal point, it is a double unless it is suffixed with `L`. 

```{r}
10.42
-8.67
10L
-8L
```

Text is stored in **character** vectors (usually called strings in other programming langauges). Characters are marked by surrounding single or double quotation marks.

```{r}
"Everything is awesome"
"1" # not the same as 1
identical(1, "1")
```

**Logical** values, or booleans, can be either true or false.

```{r}
TRUE
FALSE
```

**Factors** are a special kind of text value, where the values must be part of a predefined set of options. Factors are usually used for categorical data. 

```{r}
factor(c("correct", "correct", "incorrect"), levels = c("correct", "incorrect"))
```

Each of these kinds of values has a corresponding missing value, `NA`. These are used for missing or unrecorded data.

```{r}
NA
NA_character_
NA_integer_
NA_real_
```

## Variables and assignment

Values can be stored in **variables**. A value is placed into a variable using the assignment operator, `<-`.

```{r}
x <- 1
x
y <- -3.14
y
result <- "The results are positive"
result
```

