---
title: "Intermediate ggplot2"
author: "Jason A. Heppler and Lincoln Mullen"
---

## Aim of this worksheet

After completing this worksheet you should have a fuller grasp of the [ggplot2](http://ggplot2.org/) package. You will find the [ggplot2](http://ggplot2.org/) documentation and [R Graph Catalog](http://shiny.stat.ubc.ca/r-graph-catalog/) helpful.

Take note that there are several [ggplot2 extensions](http://www.ggplot2-exts.org/index.html) worth exploring.

You are free to attempt working through this worksheet with your own data, but we encourage you for now to use the sample datasets contained in three packages.[^1] 

## Sample datasets

Let's keep with the packages from the previous worksheet. You probably already have the data loaded, but here it is again in case you need it.

```{r, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr) # for tidy data

# install.packages(gapminder)
library(gapminder) # global data on life expectancy, GDP per capita, and population by country
data("gapminder")

# install.packages(historydata)
library(historydata) # datasets for historians
data("paulist_missions")
data("naval_promotions")
data("judges_appointments")
data("us_national_population")
data("us_state_populations")
data("early_colleges")
data("tudors")

## some subsets of data we'll try working with
california_population <- subset(us_state_populations,
                                state %in% "California")

western_state_populations <- subset(us_state_populations, 
                        state %in% c("Arizona", "California", "Colorado", "Idaho",
                                     "Montana", "Nevada", "New Mexico", "Oregon",
                                     "Utah", "Washington",  "Wyoming"))

# devtools::install_github('mdlincoln/europop')
library(europop) # https://github.com/mdlincoln/europop
data("europop")

# superfund sites CSV
superfund_sites <- read.csv("data/superfund_sites.csv") %>%
  mutate(
    Latitude = as.numeric(gsub("[^0-9.-]", "", Latitude)), 
    Longitude = as.numeric(gsub("[^0-9.-]", "", Longitude)) 
  ) 
# We'll clean up the Superfund data a bit.
keeps <- c("St", "Site.Name","Address","City","Zip","County","Latitude","Longitude","Final.Date")
superfund_sites <- superfund_sites[keeps]
superfund_sites$date <- as.Date(superfund_sites$Final.Date, "%m/%d/%y")
superfund_sites <- setNames(superfund_sites, tolower(names(superfund_sites)))
# install.packages(lubridate)
library(lubridate)
superfund_sites$year <- year(superfund_sites$date)
```

## Theming Visualizations 

Theming allows us to change *non-data* visual elements in our charts and graphics. We can take a look at how theming works in `ggplot` by looking at the source for the default `theme_gray()`.

```{r}
theme_gray <- function(base_size = 12) {
  structure(list(
    axis.line =         theme_blank(),
    axis.text.x =       theme_text(size = base_size * 0.8 , lineheight = 0.9, colour = "grey50", vjust = 1),
    axis.text.y =       theme_text(size = base_size * 0.8, lineheight = 0.9, colour = "grey50", hjust = 1),
    axis.ticks =        theme_segment(colour = "grey50"),
    axis.title.x =      theme_text(size = base_size, vjust = 0.5),
    axis.title.y =      theme_text(size = base_size, angle = 90, vjust = 0.5),
    axis.ticks.length = unit(0.15, "cm"),
    axis.ticks.margin = unit(0.1, "cm"),
 
    legend.background = theme_rect(colour="white"), 
    legend.key =        theme_rect(fill = "grey95", colour = "white"),
    legend.key.size =   unit(1.2, "lines"),
    legend.text =       theme_text(size = base_size * 0.8),
    legend.title =      theme_text(size = base_size * 0.8, face = "bold", hjust = 0),
    legend.position =   "right",
 
    panel.background =  theme_rect(fill = "grey90", colour = NA), 
    panel.border =      theme_blank(), 
    panel.grid.major =  theme_line(colour = "white"),
    panel.grid.minor =  theme_line(colour = "grey95", size = 0.25),
    panel.margin =      unit(0.25, "lines"),
 
    strip.background =  theme_rect(fill = "grey80", colour = NA), 
    strip.text.x =      theme_text(size = base_size * 0.8),
    strip.text.y =      theme_text(size = base_size * 0.8, angle = -90),
 
    plot.background =   theme_rect(colour = NA, fill = "white"),
    plot.title =        theme_text(size = base_size * 1.2),
    plot.margin =       unit(c(1, 1, 0.5, 0.5), "lines")
  ), class = "options")
}
```

(@) Create your own theme using the `theme_gray()` template above. Don't feel like you need to change everything, just try adjusting a few things like  font sizes, and so on to see what various edits do to the theme. Then, apply the theme to a chart and see how things change.

```{r}

```

### ggthemes

We can also leverage community-created themes for `ggplots` using the `ggthemes` package. 

Be sure to check out everything that [ggthemes](https://github.com/jrnold/ggthemes) adds to `ggplot`.

```{r}
ggplot(superfund_sites, aes(x = st)) +
  geom_bar() +
  theme_solarized()
```

(@) Try out a few of the provided themes in `ggthemes` with the plot above. Consult the [documentation](https://github.com/jrnold/ggthemes) to see what themes are available.

```{r}

```

## Labels and Annotation

Good visualizations need explanation, and `ggplot` gives us a way for labeling and annotating the data we visualize.

`ggplot` provides a built-in method called `labs()` for creating labels for the plot title, x- and y-axis labels, and legend labels. Let's add some labels to our chart.

```{r}
ggplot(superfund_sites, aes(x = st)) +
  geom_bar() +
  labs(
    title = "No. of Superfund Sites by State",
    x = "State",
    y = "Number of sites"
  )
```

(@) Add some informative labels to a plot you've created.

## Labels and Annotation

In addition to labeling titles and chart axis, we can label our data within the chart itself. Let's create a scatterplot of GDP and life expectancy in 1977 for the Americas and Europe.

```{r}
keepers <- subset(gap_with_colors,
                continent %in% c("Americas", "Europe") & year == 1977)

ggplot(keepers, aes(x = gdpPercap, y = lifeExp)) + 
  geom_jitter(position = position_jitter(width = 0.1, height = 0), alpha = 1/4) +
  geom_text(aes(label = country)) +
  labs(
    x = "GDP per capita",
    y = "Life expectancy"
  )
```

Note that we're calling on `geom_text` to add text labels to our chart, and we're telling `geom_text` that we want to use the `label` variable in our data frame as the text to display.

(@) The `gapminder` data frame comes with a column for setting color to the visualization. How would you apply color to the *text labels* from the data frame?

```{r}

```

(@) Try adding labels to another of the charts you've created.

```{r}

```

### ggrepel

Notice on the scatterplot above we suffer from overplotting -- the text labels overlap badly, and thus make the chart hard to interpret. We can draw on a `ggplot` extention called `ggrepel` to try and fix the problem. `ggrepel` attempts to nicely lay out labels and prevent overlap to ease readability.

```{r}
ggplot(keepers, aes(x = gdpPercap, y = lifeExp)) + 
  geom_jitter(position = position_jitter(width = 0.1, height = 0), alpha = 0.6, color="red") +
  geom_text_repel(aes(label = country), size = 3) +
  labs(
    x = "GDP per capita",
    y = "Life expectancy"
  ) +
  theme_minimal()
```

Note that we replaced the `geom_text` with `geom_text_repel`. `ggrepel` follows the same grammar of graphics for adding it to the chart. 

(@) The chart still suffers a little from overplotting. Try limiting the data to display in the chart (hint: you can subset on the data to display, e.g., countries with a GDP above or below a certain amount. See `?subset`.)

```{r}

```

## Network Analysis

`ggplot` does not come with a default way to nicely render network graphs. Luckily, we can lean on an extention to `ggplot` called `geom_net`, which provides a grammer of graphics approach to building a network visualization.

### geom_net

We will draw on the Tudor's network data to create a visualization of relationships among people. Note that although we can build network graphs using `geom_net`, we cannot do network analysis. You will need to draw on additional R libraries, [network](http://cran.rstudio.org/web/packages/network/), [sna](http://cran.rstudio.org/web/packages/sna/), and [igraph](http://cran.rstudio.org/web/packages/igraph/), to achieve that.

The `historydata` package contains several datasets that will work for network visualization: the `tudors` dataset contains spousal and parent/child relationships; the `judges_people` and `judges_appointments` can be useful for visualizing networks among people and courts. Let's start by looking at the Tudor network.

```{r}
data("tudors")
tudors %>% head(5)
```

Let's create a network of child/parent relationships and indicate those relationships by changing the color of the edges in the network. Note that this is an undirected network, but if you're working with data that needs to indicate the directionality of a connection `geom_net` provides a "directed" argument you can pass to the visualization (see `?GeomNet`).

```{r}
library(geomnet)
tudor.network <- ggplot(data = tudors, aes(from_id = person_2, to_id = person_1))
tudor.network + geom_net(label = TRUE,
           ecolour = ifelse(tudors$relationship == "child", "#a6cee3", "#b2df8a"),
           layout="fruchtermanreingold") +
  theme_net() +
  labs(title = "Tudors Parent/Child Network")
```

(@) The network can be arranged based on different network layouts available in the `sna` package. Try changing the "layout" argument to get different layouts of the network. (Hint: see `?gplot.layout`).

```{r}

```

Let's look at a slightly more complicated example, and see how we can mix in `igraph` to help achieve more nuanced network visualization. We'll turn to our `judges_appointments` dataset. We need to do a little cleaning on this first. Some of the judges in this dataset (which numbers 3,532 in total) served on some important courts and are listed separately, which means our network will create vertices where there should be only one. We can clean that up quickly:

```{r}
data("judges_appointments")
library(stringr)
# Remove appendages to court names
judges_appointments$court_name <- judges_appointments$court_name %>%
  str_replace("Chief Judge, ", "")
judges_appointments$court_name <- judges_appointments$court_name %>%
  str_replace(", Chief Judge", "")
judges_appointments$court_name <- judges_appointments$court_name %>%
  str_replace(", Chief Justice", "")
judges_appointments$court_name <- judges_appointments$court_name %>%
  str_replace("Associate Judge, ", "")
```

We're now going to take our dataframe and convert it into a [bipartite graph](http://www.scottbot.net/HIAL/index.html@p=41158.html). We will be splitting the data into a network of courts, weighted by the number of judges connecting them. 

```{r}
library(igraph)
library(ggnetwork)
courts <- graph.data.frame(judges_appointments, directed = FALSE)
V(courts)$type <- bipartite.mapping(courts)$type
courts <- bipartite.projection(courts, which = TRUE)
courts

ggnetwork(courts, layout = "fruchtermanreingold", weights = "weight")
ggnetwork(courts) %>% tail(5)
```

What we've done is created a network graph with 161 nodes representing the different courts, and 271 edges that represent our connection between courts. Many of the 3532 judges in the dataset served on multiple courts, so the edges are weighted to represent the multiple judges.

```{r}
judges.movement <- ggplot(data = courts, aes(x = x, y = y, xend = xend, yend = yend))
judges.movement + 
  geom_edges(aes(size = weight), color = "grey75") +
  geom_nodes(color = "gold", size = 2) +
  theme_minimal() +
  theme(axis.text = element_blank(),
        axis.title = element_blank(),
        panel.background = element_rect(fill = "grey25"),
        panel.grid = element_blank()) +
  labs(title = "Movement of judges between federal courts")
```

Consult the [ggnetwork](https://cran.r-project.org/web/packages/ggnetwork/vignettes/ggnetwork.html#edge-weights) documentation to see what sorts of things you can create with the package.

(@) Try adjusting the color of the vertex to correspond with the name of the court. Hint: you might have to hide the legend.

```{r}

```

## Mapping

We can leverage the `ggmap` extension to generate spatial visualizations using `ggplot`'s grammer of graphics. We can plot three types of spatial information: points, lines, and polygons. Just as we've used `ggplot` to build charts and networks, we can use it to build maps as well. 

### ggmap

Let's use our data on Superfund sites and map their occurance in the United States. We'll start by reading in our CSV file and do a little data manipulation to make it easier to work with.

```{r}
library(dplyr)
# We need to reformat the Latitude and Longitude columns using 
# the dplyr package
superfund_sites <- read.csv("data/superfund_sites.csv") %>%
  mutate(
    Latitude = as.numeric(gsub("[^0-9.-]", "", Latitude)), 
    Longitude = as.numeric(gsub("[^0-9.-]", "", Longitude)) 
  ) 
# We'll clean up the Superfund data a bit and grab only the
# rows we'll be working with and drop the rest.
keeps <- c("St", "Site.Name","Address","City","Zip","County","Latitude","Longitude","Final.Date")
superfund_sites <- superfund_sites[keeps]
# Let's fix the data information and convert it into something
# R can work with more easily.
superfund_sites$date <- as.Date(superfund_sites$Final.Date, "%m/%d/%y")
superfund_sites <- setNames(superfund_sites, tolower(names(superfund_sites)))
# Finally, we'll grab the year from the date column using the
# lubridate package and create a new column to hold that data.
library(lubridate)
superfund_sites$year <- year(superfund_sites$date)

superfund_sites %>% head(5)
```

We are, essentially, ready to create our spatial visualization. Remember that `x` is mapped to longitude and `y` is mapped to latitude. We'll also use `ggplot`'s function `coord_map()` which sets a coordinate system to something sensible for a map (it default's to the Mercator projection, but you can [change this](http://docs.ggplot2.org/current/coord_map.html)).

```{r}
ggplot(superfund_sites, aes(x = longitude, y = latitude)) +
  geom_point() +
  coord_map()
```

Our data is roughly correct, but fairly useless without a basemap underneath. There are a couple of ways we could add a basemap. First, you could use a shapefile to create a spatial dataframe that `ggplot` can map using `geom_map`. This might be handy if you are, for example, plotting information onto historical boundaries or territories. Second, we can geocode a location and use maptiles from Google, Stamen, Naver, or OpenStreetMap as our basemap. We will leverage the `get_map` function from `ggmap` to geocode the location we want, as well as specify zoom levels and map type.

```{r}
library(ggmap)
# For ease of use, we'll set the basemap object
# to a variable called basemap.
basemap <- get_map(location = "United States",
                   zoom = 4)
```

Now if we look at our basemap, we'll have a geographically-centered map of the United States with Google's roads tiles.

```{r}
ggmap(basemap)
```

We could also adjust the map tile layers if we choose (map tiles default to Google). If we wanted to use [Stamen Design](http://stamen.com)'s watercolor map, for example:

```{r}
stamen.basemap <- get_map("United States",
                   zoom = 4,
                   source = "stamen",
                   maptype = "watercolor")

ggmap(stamen.basemap)
```

Let's stick with the Google basemap. Now that we have a basemap, we can modify our `ggplot` code from above by passing the `basemap` to `ggmap` first.

```{r}
ggmap(basemap) +
  geom_point(data = superfund_sites, 
             aes(x = longitude, y = latitude),
             color = "#894D46",
             size = 0.5)
```

`ggmap` also comes with some spatial algorithmic measures. We could, for example, plot the density of points using `geom_hex()` or `geom_density2d`.

```{r}
ggmap(basemap) +
  geom_density2d(data = superfund_sites,
                 aes(x = longitude, y = latitude))
```

(@) Try the above code using `geom_hex`. What does the console tell you? What could you do to correct for this?

```{r}

```

## Create your own plots

There are a number of data sets available to you. You may try using `early_colleges`, `catholic_dioceses`, `naval_promotions`, `quasi_war`, `sarna`, `us_national_population`, or `us_state_populations` (all from the historydata package), `gapminder` (from the gapminder package), or `europop` (from the europop package).

Create three plots below, using any one or more than one of those datasets or your own. Your three plots should try to make some kind of humanistic observation. For each plot, include no more than three sentences explaining what you think the plot means. You should try to make each plot as informative as possible by using different geoms and including as many variables as is reasonable in each plot. Be sure to add good titles and labels.

You may wish to look at the [R Graph Catalog](http://shiny.stat.ubc.ca/r-graph-catalog/) to find examples of what you can do with ggplot along with sample code.

(@) Plot 1

```{r}

```

> Explanation of plot 1.

(@) Plot 2

```{r}

```

> Explanation of plot 2.

(@) Plot 3

```{r}

```

> Explanation of plot 3.


[^1]: Note a small but sometimes confusing nomenclature around these packages. In R parlance, *packages* are collections of R functions, data, and compiled code. The directory where packages are stored are called the *library*. This is somewhat counterintuitive since packages are called forth using `library()`.
