---
title: "Data Manipulation"
output: html_document
---

Load some necessary packages.

```{r, include=FALSE}
library(historydata)
library(dplyr)
library(tidyr)
library(ggplot2)
```

To work on data manipulation, we will primarily use a dataset of Methodist membership figures.

```{r}
data("methodists")
methodists 
```

# Select

We can select which columns we want by putting the name of the column in the `select()` function. Here we pick two columns.

```{r}
methodists_selected <- methodists %>% 
  select(year, members_total)

methodists %>% 
  select(year, starts_with("members"))
```

We can also get rid of columns by using the `-` sign. The `starts_with()` function is useful.

Read the documentation for this function, `?select`.

Select the `year`, `meeting`, and all the columns that begin with the word `members_`.

```{r}
methodists %>% 
  select(year, meeting, starts_with("members_"))

```

Remove the column `url`.

```{r}
methodists %>% 
  select(-url)
```

Pick just the columns that you want:

```{r}
methodists %>% 
  select(year, meeting, members_black)

```

```{r}
methodists %>% 
  select(conference, district, meeting) %>% View
```


# Filter

Filtering is more interesting. To keep certain rows, we have to pass the `filter()` function a vector of `TRUE` and `FALSE` values, one for each row. The most common way to do that is to use a comparison operator on a column of the data.

```{r}
methodists %>% 
  filter(year <= 1786 | year >= 1830) %>% View
```

```{r}
methodists %>% 
  filter(meeting == "Fairfax") %>% 
  select(year, members_total)

ggplot(methodists, aes(x = year, y = members_total, color = meeting)) + geom_point() + geom_line()
```

Can you get just the circuits with more than 1,000 members?

```{r}
methodists %>% 
  filter(members_total > 1000)
```

Can you get just the circuits in the Baltimore conference between 1820 and 1830?

```{r}
methodists %>% 
  filter(conference == "Baltimore",
         members_total < 100 | members_total > 1000)
         

```

Get just the rows from New York in 1800.

```{r}
methodists %>% 
  filter(state == "New York", year == 1800) %>% View

```

Which Methodist meetings had only black members?

```{r}
methodists %>% 
  filter(members_black == TRUE)
         

```

Bonus: some of these Methodist meetings were "missions." Load the `stringr` package and look at the documentation for `?str_detect`. Can you find which rows represent missions?

```{r}
library(stringr)
methodists %>% 
  filter(str_detect(str_to_lower(meeting), "mission")) %>% View
```


# Arrange

The `arrange()` function lets us sort. Often we want to sort a data frame by one of its columns. This can be done with the verb `arrange()`. By default `arrange()` will sort from least to greatest; we can use the function `desc()` to sort from greatest to least. In this example, we sort the data frame to get the circuits with the highest number of white members. Here are the biggest circuits in 1830. (Notice that we can create a pipeline of functions.)

```{r}
methodists %>% 
  filter(year == 1830) %>% 
  select(conference, district, meeting, members_total) %>% 
  arrange(desc(members_total)) %>% View
```

Can you find the circuits with the most African American members in a given year? Be sure to select only the necessary columns so that the results print in a meaningful way.

```{r}
methodists %>% 
  filter(year == 1786) %>% 
  select(year, meeting, members_black, members_total) %>% View
  arrange(desc(members_black)) %>%
  slice(1)
```

Which circuits had the most members overall?

```{r}
methodists %>% 
  filter(year > 1786, members_total) %>% 
  arrange(desc(members_total))
```

Which circuits had the high percentage of black members without being entirely black?

```{r}

```


# Mutate

The `mutate()` function lets us create new columns out of existing columns. The `year` column actually represents the year of the minutes, so perhaps we would like to subtract one from each year to create a new column.

```{r}
methodists %>% 
  mutate(year = year - 1) %>% View
```

Can you calculate the percentage of white members? the percentage of black members? Can you create a `members_nonwhite` column?

```{r}
methodists %>% 
  mutate(percent_white = round(members_white / members_total * 100, 2),
         percent_black = members_black / members_total * 100,
         members_nonwhite = members_total - members_white) %>% 
  select(percent_white, percent_black, everything())
```

# Left join

```{r}
cities <- c("Baltimore", "Atlanta", "Charleston")
methodist_cities <- methodists %>% 
  filter(year == 1830,
         meeting %in% cities) %>% 
  select(meeting, members_total, members_white, members_black)
methodist_cities
```

We will also create a data frame of cities with their populations in 1830 and their latitudes and longitudes. 

```{r}
data("us_cities_pop")
cities_pop <- us_cities_pop %>% 
  filter(city %in% cities, year == 1830) %>% 
  select(city, state, population, lat, lon)
cities_pop
```

It would be nice to have that data connected to our Methodists data. We can do that with `left_join()` as long as they share two columns in common.

```{r}
methodist_cities %>% 
  left_join(cities_pop, by = c("meeting" = "city"))
```

Now that you have joined the data together, can you calculate the percentage of Methodists in that city?

```{r}

```

Suppose you want to create a scatter plot faceted by state or conference. There are too many possible values for state or conference to make a useful visualization. What we really want to do is facet the plot or color the points by *region*. To do that we need to create a data frame with one row for each state, with a corresponding value for region. So for exampe, that data frame might include a value for Connecticut in the Northeast region, and Massachusetts in the Northeast region.

Create a lookup table for the Methodist data, join it to the Methodist data, and then create a visualization using ggplot2.

```{r}

```



```{r}
library(europop)
# devtools::install_github("mdlincoln/europop")

europop %>% 
  left_join(city_coords, by = c("city" = "city")) %>% View
```


# Group by and summarize

Notice that in the example above the `arrange()` function sorted the entire data frame. So when we looked for the circuits with the largest number of members, we got rows from 1825, then 1830, then 1829, then 1830, and so on. What if we wanted to get the biggest circuit from each year?

We can solve this kind of problem with what Hadley Wickham calls the "split-apply-combine" pattern of data analysis. Think of it this way. First we can *split* the big data frame into separate data frames, one for each year. Then we can *apply* our logic to get the results we want; in this case, that means sorting the data frame. We might also want to get just the top one row with the biggest number of members. Then we can *combine* those split apart data frames into a new data frame.

Observe how this works. If we want to get circuit with the most black members in each year, we can use the following code:

```{r}
methodists %>% 
  select(year, meeting, starts_with("members")) %>% 
  group_by(year) %>% 
  arrange(desc(members_black)) %>% 
  slice(5) 
```

Let's walk through that logic step by step. 

1. First we get only the years since 1786 with `filter()`, since the data does not keep track of the number of white and black members until that year.
2. Then we select only the columns that we are interested in, namely, the column for the year, the name of the meeting, and the various values for membership. We do this just so that the results print out in a useful way: in a real analysis we might decide not to throw away the other columns.
3. The crucial step is when we `group_by()` the `minutes_year`. This creates a new data-frame (the *split* step) for each unique combination of values in the variables. Notice that the printed our result says that there are 45 groups, i.e., one for each year from 1786 to 1830. (Note that you can group by combinations of columns, so, one group for each combination of city and state, for instance.)
4. Next we *apply* our logic, in this case, sorting by the column `members_colored` in descending order. This puts the rows with the biggest value at the top.
5. Next we continue to *apply* our logic with `slice()`. This function simply gives us the rows in each of the split-up data frames with that index. So `slice(1)` gives us the first row, `slice(5)` gives us this fifth row, and `slice(1:5)` gives us the first through fifth rows. 
6. The last step, *combine*, where the split-up data frames are brought back together, is done for us automatically. Note that the data frame is still grouped, however, so any subsequent data manipulation verbs will be applied to the groups rather than the whole data frame. If we wished, we could use `ungroup()`.

This particular operation, getting the top value in a split up data frame is so common that dplyr provides us with a `top_n()` function as a short cut. That function also handles ties better. (What if, for instance, two circuits both have the same biggest value?)

```{r}
methodists %>% 
  select(year, conference, meeting, members_black) %>% 
  group_by(conference) %>%
  top_n(1, members_black) %>% View
```

We get the same results more concisely and reliably, though the steps of "split-apply-combine" are perhaps somewhat less easy to see.

The data that we are currently working with has one row for each combination of a **year** and a **circuit**. We might want to know the total number of Methodists for each year. To do this, we need to group by a variable, then use an aggregation function.

```{r}
methodists_conferences <-  methodists %>% 
  filter(!is.na(conference)) %>% 
  group_by(year, conference) %>% 
  summarize(circuits = n(),
            members_total = sum(members_total))

ggplot(methodists_conferences, aes(x = year, y = members_total, group = conference)) +
  geom_line()
```

For each year, which was the biggest circuit?

```{r}
methodists %>% 
  group_by(year) %>% 
  top_n(1, members_total) %>% 
  ggplot(aes(x = year, y= members_total)) + geom_line()
```

For each year since 1786, which church had the biggest percentage of black members without being entirely black?

```{r}
methodists %>% 
  filter(year >= 1786) %>% 
  mutate(percent_black = members_black / members_total) %>%  
  filter(percent_black < 1) %>% 
  group_by(year) %>% 
  top_n(1, percent_black) %>% View

```

For the year 1825, what was the biggest meeting in each conference? In each district?

```{r}

```

For each year, what were the three biggest churches in the Baltimore conference?

```{r}

```

## Summarizing or aggregating data (`summarize()`)

In the examples using `top_n()` or `slice()` we performed a very simple kind of data summary, where we took the single row with the biggest value in a given column. This essentially boiled many rows of a data frame down into a single row. We would like to be able to summarize or aggregate a data frame in other ways as well. For instance, we often want to take the sum or the mean of a given column. We can do this using the `summarize()` function in conjunction with the `group_by()` function.

In this example, we group by the year the minutes were taken. Then we find the total number of white members for each year.

```{r}
methodists %>% 
  group_by(year) %>% 
  summarize(total_members_white = sum(members_white, na.rm = TRUE))
```

Notice that we get one row in the recombined data frame for each group in the original data frame. The value in the new column is the result of a function (in this case, `sum()`) applied to the columns in each of the split apart data frames.

There is also a special case where we might want to know how many rows were in each of the split apart (or grouped) data frames. We can use the special `n()` function to get that count. (Just like the case of `slice()` and `top_n()`, this is such a common thing to do that dplyr provides the special functions `count()` and `tally()`. You can look up their documentation to see how they work.)

```{r}
methodists %>% 
  group_by(year) %>% 
  summarize(total_meetings = n())
```

How many meetings (i.e., churches or circuits) were there in each conference in each year since 1802?

```{r}

```

What is the average number of white, black, and average number of total members for each year since 1786?

```{r}

```

Make a plot of the average percentage of black members over time.

```{r}

```

Make a plot of the average and median number of members over time.

```{r}

```

What was the average number of members in each conference for each year? Can you also make a plot of this?

```{r}

```

What was the average percentage of black members in each conference for each year? Can you also make a plot of this?

```{r}

```

## Data reshaping (`spread()` and `gather()`)

It can be helpful to think of tabular data as coming in two forms: wide data, and long data. Let's load in a table of data. This data contains total membership figures for the Virginia conference of the Methodist Episcopal Church for the years 1812 to 1830.

```{r}
library(readr)
va_methodists_wide <- read_csv("http://lincolnmullen.com/projects/worksheets/data/va-methodists-wide.csv")
va_methodists_wide
```

The first thing we can notice about this data frame is that it is very wide because it has a column for each of the years. The data is also suitable for reading because it like a table in a publication. We can read from left to right and see when certain districts begin and end and get the values for each year. The difficulties of computing on or plotting the data will also become quickly apparent. How would you make a plot of the change over time in the number of members in each district? Or how would you filter by year, or summarize by year? For that matter, what do the numbers in the table represent, since they are not given an explicit variable name?

The problem with the table is that it is not *tidy data*, because the variables are not in columns and observations in rows. One of the variables is the year, but its values are in the column headers. And another of the variables is total membership, but its values are spread across rows and columns and it is not explicitly named. 

The `gather()` function from the [tidyr](https://cran.rstudio.com/web/packages/tidyr/) package lets us turn wide data into long data. We need to tell the function two kinds of information. First we need to tell it the name of the column to create from the column headers and the name of the implicit variable in the rows. In the example below, we create to new columns `minutes_year` and `total_membership`. Then we also have to tell the function if there are any columns which should remain unchanged. In this case, the `conference` and `district` variables should remain the same, so we remove them from the gathering using the same syntax as the `select()` function.

```{r}
va_methodists_wide %>% 
  gather(minutes_year, total_membership, -conference, -district) %>% View
```

```{r}
data("dijon_prices_wide")
dijon_prices_wide %>% 
  gather(year, price, -commodity, -measure) %>% View
```


We can see the results above. There are two ways that this result is not quite what we want. Because the years were column headers they are treated as character vectors rather than integers. We can manually convert them in a later step, but we can also let `gather()` do the right thing with the `convert =` argument. Then we have a lot of `NA` values which were explicit in the wide table but which can be removed from the long table with `na.rm =`.

```{r}
age_to_bins <- function(x) {
  as.factor(ifelse(x >= 18, "adult", "child"))
}
y <- c(16, 23, 45)
age_to_bins(y)
```


```{r}
va_methodists_long <- va_methodists_wide %>% 
  gather(minutes_year, total_membership, -conference, -district, 
         convert = TRUE, na.rm = TRUE)

va_methodists_long
```

Notice that now we can use the data in ggplot2 without any problem.

```{r}
ggplot(va_methodists_long, 
       aes(x = minutes_year, y = total_membership, color = district)) +
  geom_line() +
  ggtitle("Membership of districts in the Virginia conference")
```

The inverse operation of `gather()` is `spread()`. With `spread()` we specify the name of the column which should become the new column headers (in this case `minutes_year`), and then the name of the column to fill in underneath those new column headers (in this case, `total_membership`). We can see the results below.

```{r}
# va_methodists_wide2 <- 
va_methodists_long %>% 
  spread(district, total_membership) %>% View

va_methodists_wide2
```

Just by looking at the data we can see that we got back to where we started, but we can also verify that programmatically.

```{r}
identical(va_methodists_wide, va_methodists_wide2)
```

Turning long data into wide is often useful when you want to create a tabular representation of data. (And once you have a data frame that can be a table, the `knitr::kable()` function is quite nice.) And some algorithms, such as clustering algorithms, expect wide data rather than tidy data.

For the exercise, we will use summary statistics of the number of white and black members in the Methodists by year.

```{r}
methodists_by_year_race <- methodists %>% 
  group_by(year) %>% 
  summarize(white = sum(members_white, na.rm = TRUE),
            black = sum(members_black, na.rm = TRUE),
            indian = sum(members_indian, na.rm = TRUE))
methodists_by_year_race
```

The data in `methodists_by_year_race` could be tidier still. While `white`, `black`, and `indian` are variables, it is perhaps better to think of them as two different variables. One variable would be `race`, containing the racial descriptions that the Methodists used, and another would be `members`, containing the number of members. Using the `gather()` function, create that data frame.

```{r}

```

Use the data frame you created in the previous step to create a line plot of membership over time, mapping the `race` column to the `color` aesthetic.

```{r}

```

Now use that newly tidied data frame to create a wide data frame, where the years are the column headers and the racial descriptions are the rows.

```{r}

```

Now use the same tidied data to create a wide data frame where the racial descriptions are column headers and the years are rows.

```{r}

```


## Window functions

There are a number of different kinds of window functions in R. We are going to look at two window functions, `lead()` and `lag()` which help us look for change over time. (For a fuller explanation of window functions, see the [related dplyr vignette](https://cran.rstudio.com/web/packages/dplyr/vignettes/window-functions.html).)

To understand what a window function does, it is helpful to compare it to a transformation function and an aggregation function. Suppose we have a vector with five numeric values. 

```{r}
original <- c(1.1, 2.2, 3.3, 4.4, 5.5)
```

A transformation function changes each element in the vector and returns a new value for each. In the example below, we round each element in the vector. We have a different result, but it still has five elements.

```{r}
round(original, 0)
```

In an aggregation function, we pass in a vector of numbers and get back a single value. In this case, we get the sum of the numbers.

```{r}
sum(original)
```

Aggregation functions work well with `summarize()`; transformation functions works well with `mutate()`. 

A window function gives back a vector of numbers, but a vector which has fewer useable elements than the original. It is like sliding a window over the vector. Consider the case below.

```{r}
lead(original)
lag(original)
```

The function `lead()` returns the next element of a vector in place of the original value. At the end of the vector we get an `NA` because there are no more elements left. The function `lag()` does the opposite, giving us the previous element in the vector. In that case, the first element of the returned vector is `NA`.

The `lead()` and `lag()` functions are useful for comparing one value to its previous or successor value. Suppose, for instance, that we have a vector of membership figures for each year. We can calculate the number of new members each year by subtracting the current value from its previous value.

```{r}
membership <- c(100, 150, 250, 400, 600)
membership - lag(membership)
```

Now that we understand those basics, we can apply that to the Methodist annual minutes data that we worked with in a previous lesson. Let's start by getting just the membership data from Fairfax, Virginia. 

```{r}
fairfax <- methodists %>% 
  filter(meeting == "Fairfax") %>%
  select(year, meeting, starts_with("members"), -members_indian)
fairfax
```

Now that we have the data, we can add a column for the number of new members added each year.

```{r}
fairfax %>% 
  mutate(growth = members_total - lag(members_total)) 
```

In what years did the Methodists grow in Fairfax? In what years did they shrink? Can you plot the growth?

```{r}

```

How might you use the growth figures to detect errors in the data? Keep in mind that they might have shunk or grown because of the way that they were counted.

> 

Find the growth in the number of white and black members in Fairfax, and plot them on the same chart. What is the pattern?

```{r}

```

Return back to the original `methodists` data. Beginning in 1802, the Methodists organized the data by conference, district, and meeting. For 1802 and following, calculate the growth for each conference. Which conferences were growing the most in absolute terms? Which were growing the most in relative terms (i.e., growth percentage)? Do you get a clearer picture by looking at the growth in districts? Feel free to plot the data if you wish and to add explanatory text.

```{r}

```

# Capstone

Can you aggregate the data by year and conference to compute the change percentages of African American membership? Can you make a plot of the change over time? Can you make a nice table of the data using the `knitr::kable()` function? (Hint: the Methodists didn't keep track of conferences until 1802.) (This is hard the first time you do it. You might want to start with just grouping by year, then go back and add the conferences later.)

```{r}

```

