---
title: "R Notebook"
output: html_document
---

Load the necessary spatial libraries.

```{r}
library(sp)                  # the older, more stable way
library(sf)                  # the newer, easier way
library(leaflet)             # for mapping
library(scales)              # for scales
library(tidyverse)
library(historydata)
library(USAboundaries)
data("paulist_missions")
data("catholic_dioceses")
```

## Point data

Our Paulist missions data has latitude and longitude coordinates.

```{r}
paulist_missions %>% 
  select(year, lat, long, church)
```

So does our list of Catholic dioceses.

```{r}
catholic_dioceses 
```

Let's get Paulist missions for just 1855.

```{r}
paulists_1855 <- paulist_missions %>% 
  filter(year == 1855)

dioceses_1855 <- catholic_dioceses %>% 
  filter(date <= "1855-12-31",
         event != "metropolitan")
```

## A basic leaflet map

We can create a basic leaflet map with code that is similar to a call to ggplot2. Notice that we specify a data frame, layers (cf. geoms), and in this case the aesthetics are latitude and longitude, which leaflet is smart enough to guess for us.

```{r}
leaflet(paulists_1855) %>% 
  addTiles() %>% 
  addCircleMarkers()
```

We can specify aesthetics using R's formula feature (the `~`). Aesthetics are mapped to variables in the data using a formula; properties that are not mapped to data do not use the formula. Notice also that we have moved the data frame to be specific to a single layer.

```{r}
leaflet() %>% 
  addTiles() %>% 
  addCircleMarkers(data = paulists_1855,
                   lng = ~long, lat = ~lat,
                   label = ~paste(church, city, state, sep = ", "),
                   color = "red")
```

## Exercise 1

Can you improve the display of the map above? Can you add a popup? Can you size the circles based on the number of confessions or converts? Can you add a layer of Catholic dioceses in 1855? Look at the documentation for these functions: `?addProviderTiles`, `?addCircleMarkers`. You may also need to use the `rescale()` function from the scales package.

```{r}

```

## Polygon data in sp

We can get the state boundaries of the US in 1855 using a function from the USAboundaries package.

```{r}
us_1855 <- us_states("1855-07-04")

```


We can get a quick look at what is in that spatial object using `plot()`. (Be careful doing this later. Spatial objects can get big quick, and calling plot is a good way to freeze your R session.)

```{r}
plot(us_1855)
```

This kind of data is a `SpatialPolygonsDataFrame`. We can see what is inside it by using the `str()` function. 

```{r}
str(us_1855, max.level = 2)
```

You can think of this as a set of polygons that has a data frame associated with it. We can see the associated data frame by using the `@` accessor.

```{r}
us_1855@data %>% 
  as_tibble() # so it prints nicely
```

We can notice that there are terrories, unorganized territories, and states in this data. We can't simply use dplyr functions on the data frame, since that would not change the polygons. We have to use base R functions, like this:

```{r}
us_1855_states <- us_1855[us_1855$terr_type == "State", ]
```

Note the difference.

```{r}
plot(us_1855_states)
```


## Polygon data in sf

An object that comes from the sp package is like a polygon with a data frame attached. The much newer sf package turns this relationship inside out, and gives us a data frame with polygons attached.

We can convert most objects that derived from the sp package to their equivalent sf package objects like so.

```{r}
us_1855_sf <- st_as_sf(us_1855)
```

Now we can use dplyr functions on our spatial data. This will make it vastly simpler to deal with spatial data.

```{r}
us_1855_sf %>% 
  filter(terr_type == "State") %>% 
  str(max.level = 1)
```

## Polygon data in leaflet

Plotting polygons in leaflet is similar to plotting points.

```{r}
leaflet() %>% 
  addPolygons(data = us_1855)
```

Note that leaflet can take either objects from the sp package or from the sf package.

```{r}
leaflet() %>% 
  addPolygons(data = us_1855_sf)
```

## Exercise 2

Can you take your map of Paulist missions above and include the state boundaries of the US in 1855 in an attractive way?

```{r}

```

## Reading in shapfiles

We are going to download some shapefiles of railroads in American history. These are polylines, not polygons, but the process is the same for either type.

```{r, eval = FALSE}
url <- "http://railroads.unl.edu/shared/resources/USrailshps.zip"
download.file(url = url, destfile = basename(url))
unzip(basename(url), overwrite = TRUE)
```

Shapefiles are a common format for holding spatial data. Notice that they are comprised of multiple files, all with the same name but different extensions.

```{r}
list.files(path = "USrailshps/RR1855/")
```

The most common way to load a shapefile in R is to use the rgdal package. But we are going to use the functions in sf. Note that the sf package and rgdal all call out to the same gdal/ogr library. The easiest thing to do is to point the `st_read()` function to the `.shp` file.

```{r}
rr_1855 <- st_read("USrailshps/RR1855/RR1855WGS84.shp")
```

You can plot polylines just like polygons by changing the function.

```{r}
leaflet() %>% 
  addTiles() %>% 
  addPolylines(data = rr_1855)
```


## Exercise 3

Can you add the railroads to your map of Paulist missions? What does it tell you?

```{r}

```

## Choropleth maps

So far we have only plotted data as points. But often we might want to plot data inside polygons. Let's color the status of the various states and territories in 1855. Let's count the types of territorial status that there were in 1855.

```{r}
us_1855_sf %>% 
  count(terr_type, sort = TRUE)
```

There were 31 states, 7 territories, and so on. (Note that sf is smart enough to create a new aggregated polygon for those types of territories.) It's enough that we know there are four different kinds of status that a state or territory could have.

We need to create a function. This function is going to take in the type of territory (e.g., `"State"`) and return a color. In other words the domain is a set of categories, and the range is a set of colors.

```{r}
pal_terrtype <- colorFactor(palette = topo.colors(4), domain = us_1855_sf$terr_type)
```

We can test this function. We put in types of territories, and get back colors.

```{r}
pal_terrtype(c("State", "Territory", "State"))
```

Now we can make a map that colors by territorial or state status.

```{r}
leaflet(us_1855_sf) %>% 
  addPolygons(fillColor = ~pal_terrtype(terr_type), fillOpacity = 1,
              color = "black", opacity = 1,
              label = ~full_name,
              smoothFactor = 0.2) %>% 
  addLegend("bottomright", pal = pal_terrtype, values = ~terr_type, title = "Territory type", opacity = 1)
```


## Joining spatial data

Whenever you want to plot data inside polygons in a choropleth, you need to have the data inside the same data frame that is associated with the polygons. For example, suppose we want to plot the number of converts or number of confessions per state. 

```{r}
paulists_by_state <- paulist_missions %>% 
  group_by(state) %>% 
  summarize(converts = sum(converts, na.rm = TRUE),
            confessions = sum(confessions, na.rm = TRUE),
            n = n()) %>% 
  filter(state != "Canada")
paulists_by_state
```

Since this data is no longer related to a single year, let's get a contemporary map of the United States from the USAboundaries package. We can do this by calling the same function as before, but without a data. We are also going to convert it to an sf object. As we might not always know this in advance, but Hawaii and Alaksa can be a pain, and here we can safely eliminate them. And finally our data is not in the right projection, so we are going to reproject it to [EPSG 4326](http://spatialreference.org/ref/epsg/wgs-84/), or simple latitudes and longitudes.

```{r}
us <- us_states() %>% 
  st_as_sf() %>% 
  filter(!stusps %in% c("AK", "HI")) %>% 
  st_transform(4326)
```

Now we have our geometry in one table and our data of interest in another table. How do we get them together? We do a left join. Note that the left join should always from the spatial data (left-hand side) to the data of interest (right-hand side.)

```{r}
us_with_paulists <- us %>% 
  left_join(paulists_by_state, by = c("stusps" = "state"))
```

You can View the table to see that we now have polygons and our data of interest.

```{r, eval=FALSE}
View(us_with_paulists)
```

Now we can make a map of the number of converts.

```{r}
converts_palette <- colorBin("OrRd", domain = us_with_paulists$converts,
                                bins = 7, na.color = "#808080")
leaflet(us_with_paulists) %>% 
  addTiles() %>% 
  addPolygons(fillColor = ~converts_palette(converts),
              fillOpacity = 1,
              color = "black",
              opacity = 1,
              weight = 1,
              label = ~paste(stusps, ": ", converts, sep = "")) %>% 
  addLegend("bottomright", pal = converts_palette, values = ~converts, opacity = 1, title = "Converts")
```


## Exercise 4

Can you make a map of the number of confessions? Of missions?

```{r}

```

## Exercise 5

The following code will download a shapefile from 1890 (`US_county_1890.shp` and related files) and some 1890 census materials. All of this comes from NHGIS. Look at the code book to identify some data of interest. Create a choropleth map of that data. You will need to join the spatial data to the Census data. After you've done that join, you may wish to plot only a single state to start, since that will make the functions quicker to run.

```{r, eval=FALSE}
url2 <- "http://lincolnmullen.com/projects/spatial-workshop/data/day2-afternoon.zip"
download.file(url2, destfile = basename(url2))
unzip(basename(url2), overwrite = TRUE)
unzip("day2-afternoon/US_county_1890.zip")
```

