---
title: "Creating a network"
author: "Lincoln Mullen"
date: "April 20, 2016"
output: html_document
---

```{r}
library(igraph)
library(dplyr)
library(stringr)
library(ggplot2)
library(ggforce)
library(ggraph)
library(gender)
library(humaniformat)
library(broom)
```

Downlaod the data:

```{r}
if (!file.exists("bplnetwork.txt")) {
  download.file("https://raw.githubusercontent.com/wcaleb/mining-bpl/master/bplnetwork.txt",
                "bplnetwork.txt")
}
```

Read the file:

```{r}
bpl <- read.csv("bplnetwork.txt", sep = ";")
bpl <- bpl %>% 
  mutate(source = str_replace_all(source, ",$", ""),
         target = str_replace_all(target, ",$", "")) %>% 
  filter(source != "None",
         target != "None",
         source != "Child",
         target != "Child") %>% 
  count(source, target) %>% 
  rename(weight = n)
```

Create some information about the people.

```{r}
people <- data_frame(names = c(bpl$source, bpl$target)) %>% 
  distinct() %>% 
  mutate(first_name = first_name(format_reverse(names)),
         first_name = str_replace_all(first_name, "[[:punct:]]", ""))

genders <- gender(unique(people$first_name), years = c(1800, 1860), method = "ipums") %>% 
  select(first_name = name, gender)

people <- people %>% 
  left_join(genders, by = "first_name")

to_label <- c("Garrison, William Lloyd",
              "Webb, Richard Davis",
              "Weston, Anne Warren", 
              "Phillips, Wendell",
              "Higginson, Thomas Wentworth",
              "Douglass, Frederick")
people <- people %>% 
  mutate(label = ifelse(names %in% to_label, names, NA))
```


Create the graph

```{r}
g <- graph_from_data_frame(bpl, directed = TRUE, vertices = people)
g
```

As an adjacency matrix.

```{r}
m <- as_adjacency_matrix(g, attr = "weight")
m[to_label, to_label] 
# m[to_label, to_label] %>% as.matrix %>% View
```

There are isolates which we want to remove.

```{r}
g2 <- decompose.graph(g, min.vertices = 10) 
g2 <- g2[[1]]
g2
```

A basic layout.

```{r}
ggraph(g2, layout = "igraph", algorithm = "fr") +
  geom_edge_fan(aes(edge_width = weight, edge_alpha = weight)) +
  geom_node_point(size = 1, alpha = 0.5, aes(color = gender)) +
  geom_node_text(aes(label = label)) +
  scale_edge_width("Sections borrowed", range = c(0.25, 2), guide = "none") + 
  scale_edge_alpha(range = c(0.1, 0.5), guide = "none") +
  theme_no_axes() +
  ggtitle("Abolitionist correspondence network") 
```

Calculating measures of centrality.

```{r}
betweenness(g2) %>% tidy() %>% arrange(desc(x))
eigen_centrality(g2, directed = FALSE)$vector %>% tidy() %>% arrange(desc(x))
closeness(g2) %>% tidy() %>% arrange(desc(x))
```

Create a network graph colored by different measures of centrality rather than by gender. Can you also label only the most important points programmatically, rather than listing them out?
