---
title: "Window functions"
author: ""
---

## Aims of this worksheet

This worksheet teaches you how to use window functions, a special kind of function that is useful for understanding change over time. (For a fuller explanation of window functions, see the [related dplyr vignette](https://cran.rstudio.com/web/packages/dplyr/vignettes/window-functions.html).) We can begin by loading the data and packages that we need.

```{r, message=FALSE}
library(tidyverse)
library(historydata)
data("methodists")
methodists
```

## Lag and lead

There are a number of different kinds of window functions in R. We are going to look at two window functions, `lead()` and `lag()` which help us look for change over time. 

To understand what a window function does, it is helpful to compare it to a transformation function and an aggregation function. Suppose we have a vector with five numeric values. 

```{r}
original <- c(1.1, 2.2, 3.3, 4.4, 5.5)
```

A transformation function changes each element in the vector and returns a new value for each. In the example below, we round each element in the vector. We have a different result, but it still has five elements.

```{r}
round(original, 0)
```

In an aggregation function, we pass in a vector of numbers and get back a single value. In this case, we get the sum of the numbers.

```{r}
sum(original)
```

Aggregation functions work well with `summarize()`; transformation functions works well with `mutate()`. 

A window function gives back a vector of numbers, but a vector which has fewer useable elements than the original. It is like sliding a window over the vector. Consider the case below.

```{r}
lead(original)
lag(original)
```

The function `lead()` returns the next element of a vector in place of the original value. At the end of the vector we get an `NA` because there are no more elements left. The function `lag()` does the opposite, giving us the previous element in the vector. In that case, the first element of the returned vector is `NA`.

The `lead()` and `lag()` functions are useful for comparing one value to its previous or successor value. Suppose, for instance, that we have a vector of membership figures for each year. We can calculate the number of new members each year by subtracting the current value from its previous value.

```{r}
membership <- c(100, 150, 250, 400, 600)
membership - lag(membership)
```

Now that we understand those basics, we can apply that to the Methodist annual minutes data that we worked with in a previous lesson. Let's start by getting just the membership data from Fairfax, Virginia. We will also calculate the `members_general` value for the years it is missing, and select only the columns we absolutely need.

```{r}
fairfax <- methodists %>% 
  filter(meeting == "Fairfax") %>%
  select(year, meeting, members_total, members_white, members_black)
fairfax
```

Now that we have the data, we can add a column for the number of new members added each year.

```{r}
fairfax %>% 
  mutate(growth = members_total - lag(members_total)) 
```

(@) In what years did the Methodists grow in Fairfax? In what years did they shrink? Can you plot the growth?

```{r}

```

(@) How might you use the growth figures to detect errors in the data? Keep in mind that they might have shunk or grown because of the way that they were counted.

> 

(@) Find the growth in the number of white and black members in Fairfax, and plot them on the same chart. What is the pattern?

```{r}

```

(@) Return back to the original `methodists` data. Beginning in 1802, the Methodists organized the data by conference, district, and meeting. For 1802 and following, calculate the growth for each conference. Which conferences were growing the most in absolute terms? Which were growing the most in relative terms (i.e., growth percentage)? Do you get a clearer picture by looking at the growth in districts? Feel free to plot the data if you wish and to add explanatory text.

```{r}

```

## Rolling window functions

A rolling window function is a function which computes some summary statistic such as a mean or a median over a rolling window of time. For example, given a vector of data with one point for each year, a rolling mean for each year might be calculated for the window that includes the previous five years and the following five years. This is often useful for smoothing a trend line.

We can see how this works by loading the RcppRoll package, which includes a number of these functions. Then we can use our practice vector of membership numbers. We can calculate a rolling mean with a window of three, meaning that the value for each year will be the mean of that year, the previous year, and the following year.

```{r}
library(RcppRoll)

membership
roll_mean(membership, n = 3, fill = NA)
```
Notice that we cannot compute the rolling mean for the first or last year, since they don't have a previous year or a following year so their value is undefined. When you are using a rolling window function in a call to `mutate()`, it is important to use the `fill = NA` argument to keep the vectors properly aligned.

(@) Calculate the rolling mean of the membership figures for each conference in each year.

```{r}

```

(@) Try varying the window period (`n =`) and plotting the different trends. What effect does varying the window period have?

```{r}

```

(@) Look at the package documentation for the RcppRoll package (`package?RcppRoll`). Try at least one other rolling window function besides `roll_mean()`.
